---
title: "benchModeler tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler and Modeler tutorials before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

## The benchRegressorModeler object

This object is one of the complex modelers in RMLutils. It is a multivariate modeler in which a certain linear constraint (to be specified by the user) must be satisfied between the outputs of its internal modelers.

```{r}
# Generate an appropriate dataset for benchModeler
original_data.dt <- generateRegressionData(n_samples = 1000, generate_random_weights = FALSE)

# Define new variables in the data that have a specific constraint
# In our case: regression_target = A + B
data.dt <- data.table::copy(original_data.dt)
data.dt[, A := round(runif(dim(data.dt)[1], min = 0, max = 5))]
data.dt[, B := regression_target - A]

# Define train and test partitions
set.seed(273)
data.keys <- 1:nrow(data.dt)
train_size = round(0.8 * nrow(data.dt))
smplr <- srsSampler(key = data.keys, ssize = train_size)
train.keys <- getSample(smplr)
test.keys <- setdiff(data.keys, train.keys)
```

### Instalación de osqp, gsubfn y Matrix

First, we need to check if the `osqp`, `gsubfn` and `Matrix` libraries are installed:

```{r}
require("osqp")
require("gsubfn")
require("Matrix")
```

If they are not, we can easily install them from CRAN, using the command `install.packages(...)`.

The `osqp` library version must be at least `1.0.0`.

### Instantiation

The benchRegressorModeler requires the following parameters to be instantiated:

- `submodels`: List of modelers to predict their target variables.
- `benchmarks`: String or vector of strings indicating the linear constraints that the benchModeler must satisfy.
- `optWeights`: String indicating the type of optional weights, which can be `"uniform"` or `"MSE"`. `"uniform"` is taken by default.
- `models`: (Optional) List of modelers to use. Only those whose target variables are within benchmarks will be selected. If this parameter is provided, the `submodels` parameter must not be provided at the same time.

In this tutorial, ranger modelers will be used to define the benchModeler. The user should check that the `ranger` library is installed.

We first define the submodelers:

```{r}
# Define the parameters of the submodelers
# Note that none of the modelers see the target variable of another. They only see their own
params_b1 <- list(
  "num.trees" = 20,
  "max.depth" = 20
)
target_b1 <- "regression_target"
regressors_b1 <- setdiff(colnames(original_data.dt), target_b1)

params_b2 <- list(
  "num.trees" = 30,
  "max.depth" = 10
)
target_b2 <- "A"
regressors_b2 <- setdiff(colnames(original_data.dt), c(target_b2, target_b1))

params_b3 <- list(
  "num.trees" = 40,
  "max.depth" = 15
)
target_b3 <- "B"
regressors_b3 <- setdiff(colnames(original_data.dt), c(target_b3, target_b1))

modeler1 <- rangerRegressorModeler(params = params_b1, target = target_b1, regressors = regressors_b1)

modeler2 <- rangerRegressorModeler(params = params_b2, target = target_b2, regressors = regressors_b2)

modeler3 <- rangerRegressorModeler(params = params_b3, target = target_b3, regressors = regressors_b3)
```

We define the benchModeler:

```{r}
b_modeler <- benchRegressorModeler(submodels = list(modeler1, modeler2, modeler3),
                                   benchmarks = "A+B-regression_target==0")
```

### crossValidate

```{r}
# Define a cost function
cf <- r2CostFunction()
crossValidate(modeler = b_modeler,
              data = data.dt,
              CVSampler = cvSampler(key = data.keys, folds = 3),
              costFunction = cf)
```

### train

```{r}
b_modeler <- train(b_modeler,
                   data = data.dt,
                   subSet = train.keys)
```

### getPredictions

```{r}
preds <- getPredictions(object = b_modeler,
                        data = data.dt,
                        subSet = test.keys)
head(preds)
```

### evaluate

```{r}
# Define a cost function
cf <- r2CostFunction()
# Evaluation on the training set
evaluate(object = b_modeler,
         data = data.dt,
         cost_function = cf,
         subSet = train.keys)
# Evaluation on the test set
evaluate(object = b_modeler,
         data = data.dt,
         cost_function = cf,
         subSet = test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "bench_reg.mlu"
modelSave(object = b_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = data.dt,
         cost_function = cf,
         subSet = test.keys)
```

