---
title: "costFunction tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler and Modeler tutorials before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

## Use and types of cost functions

Cost functions are the main instrument in RMLutils for evaluating modelers. Their functionality is integrated with the modelers themselves and can be used directly through their `evaluate` or `crossValidate` methods or called manually, specifying the modeler to evaluate or the metrics needed to calculate the cost.

All cost functions have the following methods:

- `getCost`: Calculates the cost from specified data and a modeler.

- `getReducedCost`: Calculates the averaged cost for different k-folds. It is used automatically when using the `crossValidate` method of any modeler. It should be noted that the averaging is not a simple weighted mean, but is calculated taking into account the non-linear elements that exist in the operations to calculate each individual score, so the metadata obtained from each k-fold is used to compute the final score.

Additionally, getCost returns a dictionary of key metrics for the cost calculation that may be of interest to the user.

The outputs are named lists, structured first at the target variable level, and within each one, another list containing the cost and other metrics.

To use the `getCost` method, the following parameters must be provided:

- `object`: Cost function to use.

- `data`: Dataset on which to obtain the cost.

- `modeler`: Modeler from which to obtain the predictions.

- `fSample`: Full sample (the keys of the complete dataset). Some cost functions do not need this parameter.

- `cSample`: Complementary sample. In general, the keys of the data on which we want to obtain the metric/score, for example, those of the training set or the test set.

Since the getReducedCost() method is already used automatically in cross-validation, it will not be developed further in this tutorial, as the general user should not need to access it manually to obtain results.

On the other hand, the `evaluate` method provides a similar interface to getCost, and is designed to provide only the cost (without additional metrics). It has the following parameters:

- `object`: Modeler from which to obtain the predictions.

- `data`: Dataset on which to obtain the cost.

- `cost_function`: Cost function or list of cost functions, to evaluate the internal modelers of complex modelers, such as prePredModeler and classifRegModeler.

- `subSet`: Vector of indices to select a subset of the data.

- `cSample`: (optional) Argument for compatibility with some particular cost functions. In general, the user should use subSet to define the subset of data on which to evaluate the modeler.

The user is free to choose the method for evaluating their modelers.

## Cost functions for classification

The following types exist:

- `confusionMatrixCostFunction`
- `accuracyCostFunction`
- `precisionCostFunction`
- `recallCostFunction`
- `f1ScoreCostFunction`
- `ROCAUCCostFunction`

Except for `ROCAUCCostFunction`, all classification cost functions use ConfusionMatrixCostFunction as an intermediary to calculate their cost, so their output always contains a confusion matrix.

We define a dataset and a classifierModeler for these cost functions.

```{r}
# Define common data for the classification task
c_data.dt <- generateMulticlassClassificationData(n_samples = 1000,
                                                  generate_random_weights = FALSE)
# Define train/test partitions
set.seed(273)
c_data.keys <- 1:nrow(c_data.dt)
c_train_size = round(0.8 * nrow(c_data.dt))
c_smplr <- srsSampler(key = c_data.keys, ssize = c_train_size)
c_train.keys <- getSample(c_smplr)
c_test.keys <- setdiff(c_data.keys, c_train.keys)
head(c_data.dt)

# Define model parameters
c_target <- "classification_target"
c_regressors <- setdiff(colnames(c_data.dt), c_target)
c_params <- list(
  "num.trees" = 10,
  "max.depth" = 2
)
# Instantiate the model
classifier <- rangerClassifierModeler(params = c_params,
                                      target = c_target,
                                      regressors = c_regressors)
# And train it
classifier <- train(classifier, data = c_data.dt, subSet = c_train.keys)
```

### confusionMatrixCostFunction

Cost function that returns the confusion matrix for a classification modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `class_map`: classMapping object to use. maxScoreClass by default.
- `normalize`: String indicating how to normalize the matrix. Possible values are:

  - `"true"`: Normalization by true classes (rows).
  - `"pred"`: Normalization by predicted classes (columns).
  - `"all"`: Normalization by total classes.
  - `NULL`: No normalization (default).

- `class_names`: Vector of strings with the levels of the categorical target variable.

```{r}
# Example of use
cf <- confusionMatrixCostFunction(weightsColumn = "",
                                  class_map = maxScoreClass(),
                                  normalize = NULL,
                                  class_names = levels(c_data.dt[["classification_target"]]))
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### accuracyCostFunction

Cost function that returns accuracy $(TP+TN)/(TP+FP+TN+FN)$ for a classification modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `class_map`: classMapping object to use. `maxScoreClass` by default.
- `average`: String indicating the averaging method (in general it will only affect the multiclass case). Possible values are:

  - `"micro"`: Global averaged metric (hit vs miss), weighted by sampling weights, if any.
  - `"macro"`: Stratified average metric by class, weighted by sampling weights, if any.
  - `"weighted"`: Stratified average metric by class, weighted by sampling weights, if any, and by the support of each class.

```{r}
# Example of use
cf <- accuracyCostFunction(weightsColumn = "",
                           class_map = maxScoreClass(),
                           average = "micro")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### precisionCostFunction

Cost function that returns precision $(TP)/(TP+FP)$ for a classification modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `class_map`: classMapping object to use. `maxScoreClass` by default.
- `average`: String indicating the averaging method (in general it will only affect the multiclass case). Possible values are:

  - `"micro"`: Global averaged metric (hit vs miss), weighted by sampling weights, if any.
  - `"macro"`: Stratified average metric by class, weighted by sampling weights, if any.
  - `"weighted"`: Stratified average metric by class, weighted by sampling weights, if any, and by the support of each class.

```{r}
# Example of use
cf <- precisionCostFunction(weightsColumn = "",
                            class_map = maxScoreClass(),
                            average = "micro")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### recallCostFunction

Cost function that returns recall/sensitivity or true positive rate $(TP)/(TP+FN)$ for a classification modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `class_map`: classMapping object to use. `maxScoreClass` by default.
- `average`: String indicating the averaging method (in general it will only affect the multiclass case). Possible values are:

  - `"micro"`: Global averaged metric (hit vs miss), weighted by sampling weights, if any.
  - `"macro"`: Stratified average metric by class, weighted by sampling weights, if any.
  - `"weighted"`: Stratified average metric by class, weighted by sampling weights, if any, and by the support of each class.

```{r}
# Example of use
cf <- recallCostFunction(weightsColumn = "",
                            class_map = maxScoreClass(),
                            average = "micro")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### f1ScoreCostFunction

Cost function that returns the F1-score for a classification modeler and specified data. It is the harmonic mean of precision and recall/sensitivity.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `class_map`: classMapping object to use. `maxScoreClass` by default.
- `average`: String indicating the averaging method (in general it will only affect the multiclass case). Possible values are:

  - `"micro"`: Global averaged metric (hit vs miss), weighted by sampling weights, if any.
  - `"macro"`: Stratified average metric by class, weighted by sampling weights, if any.
  - `"weighted"`: Stratified average metric by class, weighted by sampling weights, if any, and by the support of each class.
  
```{r}
# Example of use
cf <- f1ScoreCostFunction(weightsColumn = "",
                            class_map = maxScoreClass(),
                            average = "micro")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### ROCAUCCostFunction

Cost function that returns the area under the ROC (Receiver Operating Characteristic) curve for a classification modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `average`: String indicating the averaging method (in general it will only affect the multiclass case). Possible values are:

  - `"micro"`: Global averaged metric (hit vs miss), weighted by sampling weights, if any.
  - `"macro"`: Stratified average metric by class, weighted by sampling weights, if any.
  - `"weighted"`: Stratified average metric by class, weighted by sampling weights, if any, and by the support of each class.

```{r}
# Example of use
cf <- ROCAUCCostFunction(weightsColumn = "",
                         average = "micro")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)

# We can plot the ROC curve with the output of getCost, since it returns TPR and FPR:
tpr <- cost$classification_target$class_metrics$tpr
fpr <- cost$classification_target$class_metrics$fpr
plot(fpr, tpr, type = "l")

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

If we select a different `average`, we can obtain ROC curve results by category, which can be useful.

```{r}
# Example of use
cf <- ROCAUCCostFunction(weightsColumn = "",
                         average = "weighted")

# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf, data = c_data.dt, modeler = classifier, cSample = c_test.keys)

# We can plot the ROC curve for a particular category, such as Target_3, with the output of getCost, since it returns TPR and FPR:
tpr <- cost$classification_target$class_metrics$Target_3$tpr
fpr <- cost$classification_target$class_metrics$Target_3$fpr
plot(fpr, tpr, type = "l")

print(paste0("AUC for category Target_3: ", cost$classification_target$class_metrics$Target_3$auc))

# Evaluation with evaluate (returns only the cost)
evaluate(object = classifier,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

## Cost functions for regression

The following types exist:

- `r2CostFunction`
- `squareBiasCostFunction`
- `MSECostFunction`
- `groupMSECostFunction`
- `MCSubRBMSECostFunction`
- `holdoutTemporalCostFunction`

We define a dataset and a regressorModeler for these cost functions.

```{r}
# Define common data for the regression task
r_data.dt <- generateStratifiedDomainModelAssistedEstimateData(n_samples = 1000,
                                                               generate_random_weights = TRUE)
# Define train/test partitions
set.seed(273)
r_data.keys <- 1:nrow(r_data.dt)
r_train_size = round(0.8 * nrow(r_data.dt))
r_smplr <- srsSampler(key = r_data.keys, ssize = r_train_size)
r_train.keys <- getSample(r_smplr)
r_test.keys <- setdiff(r_data.keys, r_train.keys)
head(r_data.dt)

# Define model parameters
r_target <- "target"
r_regressors <- setdiff(colnames(r_data.dt), r_target)
r_params <- list(
  "num.trees" = 20,
  "max.depth" = 10
)
# Instantiate the model
regressor <- rangerRegressorModeler(params = r_params,
                                    target = r_target,
                                    regressors = r_regressors)
# And train it
regressor <- train(regressor, data = r_data.dt, subSet = r_train.keys)
```

### r2CostFunction

Cost function that returns the $R^{2}$ metric for a regression modeler and specified data. The `fSample` parameter indicating the keys of the complete dataset must be specified.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).

```{r}
# Example of use
cf <- r2CostFunction(weightsColumn = "weights")
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf,
                data = r_data.dt,
                modeler = regressor,
                fSample = r_data.keys,
                cSample = r_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = regressor,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### squareBiasCostFunction

Cost function that returns the squared bias of the predictions for a regression modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).

```{r}
# Example of use
cf <- squareBiasCostFunction(weightsColumn = "weights")
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf,
                data = r_data.dt,
                modeler = regressor,
                cSample = r_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = regressor,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### MSECostFunction

Cost function that returns the mean squared error of the predictions for a regression modeler and specified data.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).

```{r}
# Example of use
cf <- MSECostFunction(weightsColumn = "weights")
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf,
                data = r_data.dt,
                modeler = regressor,
                cSample = r_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = regressor,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### groupMSECostFunction

Cost function that returns the weighted mean squared error of a function for each specified group/stratum. That is, after prediction, each group provides a unique metric such as sum or mean, and the mean squared error is calculated by comparing this sum/mean of predictions and original values.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `groupWeightsFunction`: Averaging/combining function for the weights of each group.
- `groupFunction`: Function to average/combine the values of the target variable for a complete group.
- `groupColumn`: Name of the column that defines the group/stratum/domain in the dataset.

```{r}
# Example of use
cf <- groupMSECostFunction(weightsColumn = "weights",
                           groupWeightsFunction = sum, 
                           groupColumn = "strata_class", 
                           groupFunction = mean)
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf,
                data = r_data.dt,
                modeler = regressor,
                cSample = r_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = regressor,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### groupTotalErrorCostFunction

A cost function that calculates the weighted absolute error by group, for each group element, and then combines it to obtain the total absolute prediction error.

Parameters:

- `weightsColumn`: Name of the column with sampling weights in the data (if it exists).
- `groupVars`: String or string vector indicating the variables by which to group the data.

```{r}
# Example of use
cf <- groupTotalErrorCostFunction(weightsColumn = "weights",
                                  groupVars = c("strata_class", "domain_class"))
# Evaluation with getCost (returns metadata)
cost <- getCost(object = cf,
                data = r_data.dt,
                modeler = regressor,
                cSample = r_test.keys)
print(cost)

# Evaluation with evaluate (returns only the cost)
evaluate(object = regressor,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### MCSubRBMSECostFunction

A cost function that outputs the estimator, bias MSE and Monte Carlo error of
the estimator based on Zhang, L.-C., Sanguiao-Sande, L., & Lee, D.
(2025). Design-Based Predictive Inference. Journal of Official Statistics,
41(1), 404-432. \url{https://journals.sagepub.com/doi/10.1177/0282423X241277719}.

Since this method requires performing a series of sub-samplings, it is recommended to use it directly with crossValidate (which calls getReducedCost).

Parameters:

- `sampler`: Sampler used to obtain the sample.
- `sampleKeys`: Set of keys used to define the training set for the model.
- `nonSampleData`: Data complementary to the training set.

```{r}
# Example of use
cf <- MCSubRBMSECostFunction(sampler = r_smplr,
                             sampleKeys = r_train.keys, 
                             nonSampleData = r_data.dt[-r_train.keys])
# Evaluation with getCost (returns metadata)
crossValidate(regressor,
              r_data.dt[r_train.keys,],
              srsSampler(1:length(r_train.keys),150), 
              cf,
              times = 100)# 100 sub-samples
```

### holdoutTemporalCostFunction

Complex cost function that combines another type of cost function with a cvTemporalSampler. Useful when intending to evaluate predictions where the data has some temporal dependence.

Parameters:

- `costFunction`: Cost function to use.
- `cvTemporalSampler`: cvTemporalSampler to use.

First, we generate a dataset with temporal dependence and a cvTemporalSampler:

```{r}
# Define a dataset with 36 periods (months)
years <- c(2023, 2024, 2025)
total_periods <- c(sapply(years, function(y){
  months <- 1:12
  months[1:9] <- paste0("0", months[1:9])
  return(paste0("MM", months, y))
}, USE.NAMES = FALSE))

# Generate data for each period
data.dt <- rbindlist(lapply(total_periods, function(p){
  dt <- generateRegressionData(n_samples = 100,
                               generate_random_weights = FALSE)
  dt[, period := p]
  return(dt)
}))
head(data.dt)

smplr <- cvTemporalSampler(key = 1:nrow(data.dt),
                           data = data.dt,
                           identifier = "period",
                           folds = 5,
                           n_valPeriods = 6, # Require 6 months for validation
                           orderPeriods = total_periods)
```

Then, we define a modeler for this data and train it:

```{r}
# Define modeler and train it
params <- list(
  "num.trees" = 50,
  "max.depth" = 20
)
target <- "regression_target"
regressors <- setdiff(colnames(data.dt), target)

modeler <- rangerRegressorModeler(params = params,
                                  target = target,
                                  regressors = regressors)

modeler <- train(modeler, data = data.dt)
```

Finally, we define the complete cost function and obtain the cost:

```{r}
cf <- holdoutTemporalCostFunction(costFunction = r2CostFunction(),
                                  cvTemporalSampler = smplr)

cost <- getCost(object = cf,
                data = data.dt,
                modeler = modeler,
                fSample = 1:nrow(data.dt),
                cSample = 1:nrow(data.dt))
print(cost$regression_target$cost)
```

