---
title: "H2O modeler tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler and Modeler tutorials before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

**Warning: Due to the use of the H2O cloud, this document cannot be compiled into an HTML file. To see the results, the user can manually run the code chunks in order.**

## Initializing the H2O cloud

The big difference between h2o and the other machine learning libraries in RMLutils is that operations do not happen in R, but in the h2o cloud. Previously, we must initialize the cloud and convert our data to h2o dataframes before calling any modeler object.

First, we check if the `h2o` library is installed:

```{r}
require("h2o")
```

If it is not, we can easily install it from CRAN, using the command `install.packages("h2o")`.

The operation of the h2o cloud can be problematic depending on the version of java on our computer or server.

We propose two alternatives to the user for its operation. If neither works, they should refer to the official h2o and java documentation to try to solve it in their specific case.

### Directly through the h2o library itself

If we have a compatible version of java installed, we simply use the command `h2o.init(port = 50000, startH2O = TRUE)` from the h2o library (changing the port if 50000 is already in use). If this does not work, we try the other option.

### Manual creation of the h2o cloud

The user must download a compatible version of h2o on their own machine (checking that it is a file called h2o.jar) and must run a command from the command line to launch an h2o cloud on a port of their choice. Obviously, the user should have a compatible version of java downloaded, for example, version 1.8.0.

On a linux server, for example, the command would be:

```
"/path/to/java" -jar "path/to/h2o.jar" -port 50000 &
```

"&" makes the process run in the background and does not block the command line. It is especially useful if we want to stop the process or run any other command while the cloud is running.

After this, in the script we are running, we must call `h2o.init()`, but indicating that it should not start the h2o cloud, since we have already activated it, and specifying the port where it is:

```
h2o.init(port = 50000, startH2O = FALSE)
```

## Using the H2O cloud

H2O modelers work with data.tables and H2OFrames interchangeably. The user is free to use whichever method they find more convenient.

When passing objects to the h2o cloud, it is recommended to assign them keys that represent what they are, since if we later want to examine the contents of the cloud, it is much more intuitive to see variable names as keys instead of the keys automatically generated by h2o.

```{r}
# Start the H2O cloud (through the library itself)
h2o.init(port = 54321, startH2O = TRUE)

# Generate some datasets (data.table) and pass them to the cloud
r_data.dt <- generateRegressionData(n_samples = 100, generate_random_weights = FALSE)
c_data.dt <- generateBinaryClassificationData(n_samples = 100, generate_random_weights = FALSE)

r_data.h2o <- h2o::as.h2o(r_data.dt, destination_frame = "reg_data")
c_data.h2o <- h2o::as.h2o(c_data.dt, destination_frame = "classif_data")

# See the contents of the cloud
print(h2o.ls())

# Delete the classification data
h2o.rm("classif_data")

# Check that it is no longer in the cloud
print(h2o.ls())

# See the column types of reg_data
print(h2o.getTypes(r_data.h2o))

# Shut down the cloud
#h2o.shutdown(prompt = FALSE)
```

In some cases, clouds may hang, occupying RAM memory on our computer or server. To close them, we must delete them from the command line. To do this, we scan the processes we have open and manually kill the h2o process.

On linux (feel free to use any other regular expression pattern to search for processes):

```
ps aux | grep <username>
```

Once its identifier is located (for example, 1173431), we kill it:

```
kill -9 <process_id>
```

On windows, H2O clouds can also be found and deleted in the task manager (they appear as a java process).

## H2OClassifierModeler

H2O modelers work with both data in data.table format and H2OFrame format. The second is slightly more efficient because it does not require constantly uploading data for any operation, but it requires the user to actively manage the H2O cloud.
For the classification model, we will upload the data to the H2O cloud.

```{r}
# Define common data for the classification task
c_data.dt <- generateBinaryClassificationData(n_samples = 1000,
                                              generate_random_weights = FALSE)
# Define train/test partitions
set.seed(273)
c_data.keys <- 1:nrow(c_data.dt)
c_train_size = round(0.8 * nrow(c_data.dt))
c_smplr <- srsSampler(key = c_data.keys, ssize = c_train_size)
c_train.keys <- getSample(c_smplr)
c_test.keys <- setdiff(c_data.keys, c_train.keys)
head(c_data.dt)

# Start the cloud
#h2o.init(port = 50000, startH2O = TRUE)

c_data.h2o <- h2o::as.h2o(c_data.dt, destination_frame = "classif_data")
```

### Instantiation

The H2OClassifierModeler requires the following parameters to be instantiated:

- `model`: Name of the H2O model to use. The following are available:

  - `adaBoost` (Binary classification)
  - `decision_tree` (Binary classification)
  - `deeplearning`
  - `gam` (Generalized additive model)
  - `gbm` (Gradient boosting model)
  - `glm` (Generalized linear model)
  - `naiveBayes`
  - `svm` (Support vector machine, Binary classification)
  - `randomForest`
  - `rulefit`
  - `xgboost` (Only works on linux machines)

- `params`: list with the names of the parameters required for the chosen model. It is recommended to consult the H2O documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

```{r}
# Define model parameters
target <- "classification_target"
regressors <- setdiff(colnames(c_data.h2o), target)
params <- list() # Will use predefined parameters
h2o_cls_modeler <- H2OClassifierModeler(model = "randomForest",
                                        params = params,
                                        target = target,
                                        regressors = regressors)
```

### crossValidate

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = h2o_cls_modeler,
              data = c_data.h2o, # Complete data
              CVSampler = cvSampler(key = c_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

### train

```{r}
h2o_cls_modeler <- train(object = h2o_cls_modeler, data = c_data.h2o, subSet = c_train.keys)
```

### getPredictions

When predicting, classifier modelers return a matrix of scores:

```{r}
preds <- getPredictions(object = h2o_cls_modeler,
                        data = c_data.h2o,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

A particular class mapping can be specified for this object, of which there are:

- `maxScoreClass()`: (default), chooses the class with the highest score.
- `binaryThresholdClass()`: Only for binary target. Allows specifying a selection threshold for the positive class.

```{r}
clss <- getClasses(object = h2o_cls_modeler,
                   data = c_data.h2o,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)

clss <- getClasses(object = h2o_cls_modeler,
                   data = c_data.h2o,
                   subSet = c_test.keys,
                   class_mapping = binaryThresholdClass(threshold = 0.65,
                                                        positive_class = "positive"))
head(clss)
```

### evaluate

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = h2o_cls_modeler,
         data = c_data.h2o,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = h2o_cls_modeler,
         data = c_data.h2o,
         cost_function = cf,
         subSet = c_test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "h2o_cls.mlu"
modelSave(object = h2o_cls_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.h2o,
         cost_function = cf,
         subSet = c_test.keys)

# Shut down the H2O cloud
#h2o.shutdown(prompt = FALSE)
```

**Important: If the H2O cloud is closed before saving the model, we will not be able to recover it, because all its contents will be deleted.**

## H2ORegressorModeler

H2O modelers work with both data in data.table format and H2OFrame format. The second is slightly more efficient because it does not require constantly uploading data for any operation, but it requires the user to actively manage the H2O cloud.
For the regression model, we will not upload the data to the H2O cloud, we will let the modeler itself do it automatically.

```{r}
# Define common data for the regression task
r_data.dt <- generateRegressionData(n_samples = 1000,
                                    generate_random_weights = FALSE)
# Define train/test partitions
set.seed(273)
r_data.keys <- 1:nrow(r_data.dt)
r_train_size = round(0.8 * nrow(r_data.dt))
r_smplr <- srsSampler(key = r_data.keys, ssize = r_train_size)
r_train.keys <- getSample(r_smplr)
r_test.keys <- setdiff(r_data.keys, r_train.keys)
head(r_data.dt)

# Start the cloud
#h2o.init(port = 54323, startH2O = TRUE)
```

### Instantiation

The H2ORegressorModeler requires the following parameters to be instantiated:

- `model`: Name of the H2O model to use. The following are available:

  - `deeplearning`
  - `gam` (Generalized additive model)
  - `gbm` (Gradient boosting model)
  - `glm` (Generalized linear model)
  - `randomForest`
  - `rulefit`
  - `xgboost` (Only works on linux machines)

- `params`: list with the names of the parameters required for the chosen model. It is recommended to consult the H2O documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the regression task.

```{r}
# Define model parameters
target <- "regression_target"
regressors <- setdiff(colnames(r_data.dt), target)
params <- list() # Will use predefined parameters
h2o_reg_modeler <- H2ORegressorModeler(model = "deeplearning",
                                       params = params,
                                       target = target,
                                       regressors = regressors)
```

### crossValidate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = h2o_reg_modeler,
              data = r_data.dt, # Complete data
              CVSampler = cvSampler(key = r_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

### train

```{r}
h2o_reg_modeler <- train(object = h2o_reg_modeler, data = r_data.dt, subSet = r_train.keys)
```

### getPredictions

When predicting, regressor modelers return a matrix with the predictions:

```{r}
preds <- getPredictions(object = h2o_reg_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = h2o_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = h2o_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "h2o_reg.mlu"
modelSave(object = h2o_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)

# Shut down the H2O cloud
h2o.shutdown(prompt = FALSE)
```

**Important: If the H2O cloud is closed before saving the model, we will not be able to recover it, because all its contents will be deleted.**
