---
title: "Modeler objects tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler tutorial before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

This document will only cover simple modelers, except for H2O ones,
which will be explained in another tutorial. Likewise, the other complex modelers
each have an associated tutorial.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

## Structure, types and methods of modelers

Modelers are the main object of the MLUtils libraries, since they are
responsible for training and predicting with models from different libraries
(LightGBM, H2O, etc.). Their names follow the structure of
model + task + "modeler", with the following types:

- Regression:
    - Simple modelers
      - `directRegressorModeler`
      - `HTRegressorModeler`
      - `strataHTRegressorModeler`
      - `strataMeanRegressorModeler`
      - `H2ORegressorModeler`
      - `LGBRegressorModeler`
      - `rangerRegressorModeler`
      - `catboostRegressorModeler`
      - `miceRegressorModeler`
    - Complex modelers
      - `classifRegRegressorModeler`
      - `benchRegressorModeler`
      - `preproRegressorModeler`
      - `prePredRegressorModeler`
      - `memoryRegressorModeler`
- Classification
    - Simple modelers
      - `H2OClassifierModeler`
      - `LGBClassifierModeler`
      - `rangerClassifierModeler`
      - `catboostClassifierModeler`
      - `miceClassifierModeler`
    - Complex modelers
      - `preproClassifierModeler`
      
All modelers have a common set of methods:

- `train`: Trains a modeler.
- `crossValidate`: Applies cross-validation to a modeler to obtain a metric of its performance defined from a cost function, but does not train it. (For more information on cost functions, it is recommended to review their associated tutorial).
- `getPredictions`: Obtains the predictions of the modeler for a given dataset. (The predictions of classifier modelers are a matriz of probabilities/scores, not the classes themselves).
- `modelSave`: Saves the model to a compressed file.
- `modelLoad`: Loads a model from a file.
- `evaluate`: Evaluates a trained model for a specific cost function.

And classifiers have one additional one:

- `getClasses`: Provides the predicted classes by a trained model, for
a given dataset.

Modelers have an analogous operation, regardless of their type.
In the rest of this tutorial, most simple modelers will be developed with
usage examples.      
      
## Classification modelers

```{r}
# Define common data for the classification task
c_data.dt <- generateBinaryClassificationData(n_samples = 1000,
                                              generate_random_weights = FALSE)
# Define train/test partitions
set.seed(273)
c_data.keys <- 1:nrow(c_data.dt)
c_train_size = round(0.8 * nrow(c_data.dt))
c_smplr <- srsSampler(key = c_data.keys, ssize = c_train_size)
c_train.keys <- getSample(c_smplr)
c_test.keys <- setdiff(c_data.keys, c_train.keys)
head(c_data.dt)
```

### LGBClassifierModeler

Implementation of LightGBM in RMLutils for classification tasks (lightweight gradient boosting framework with tree-based models).

#### Installation of lightgbm
First, we must check if the lightgbm library is installed:

```{r}
require("lightgbm")
```

If it is not, we can easily install it from CRAN, using the command `install.packages("lightgbm")`.

#### Instantiation

The LGBClassifierModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `lgb.train.` It is recommended to consult the lightgbm documentation for more information.
In the case of classification, the `objective` parameter can be `"multiclass"` (which works for both multiclass and binary target) or `"binary"` (only for binary target).

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

- `positive_class`: (Optional) For the case of `objective = "binary"`, it is recommended to specify the name or identifier of the positive class, otherwise the order according to the factor levels will be assumed, with the first level being the negative class and the second the positive one.

```{r}
# Define model parameters
target <- "classification_target"
regressors <- setdiff(colnames(c_data.dt), target)
params <- list(
  "learning_rate" = 0.07,
  "nrounds" = 10,
  "objective" = "binary"
)
lgb_cls_modeler <- LGBClassifierModeler(params = params,
                                        target = target,
                                        regressors = regressors,
                                        positive_class = "positive")

# Alternatively:
# params <- list(
#   "learning_rate" = 0.07,
#   "nrounds" = 10,
#   "objective" = "multiclass"
# )
# lgb_cls_modeler <- LGBClassifierModeler(params = params,
#                                         target = target,
#                                         regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = lgb_cls_modeler,
              data = c_data.dt, # Complete data
              CVSampler = cvSampler(key = c_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

In this case, for the variable `"classification_target"` perfect accuracy is obtained.

#### train

After determining a suitable set of parameters with crossValidate, we can train the chosen model.

```{r}
lgb_cls_modeler <- train(object = lgb_cls_modeler,
                         data = c_data.dt,
                         subSet = c_train.keys)

# Alternatively:
# lgb_cls_modeler <- train(object = lgb_cls_modeler,
#                          data = c_data.dt[c_train.keys],
```

#### getPredictions

When predicting, classifier modelers return a matrix of scores:

```{r}
preds <- getPredictions(object = lgb_cls_modeler,
                        data = c_data.dt,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

A particular class mapping can be specified for this object, of which there are:

- `maxScoreClass()`: (default), chooses the class with the highest score.

- `binaryThresholdClass()`: Only for binary target. Allows specifying a selection threshold for the positive class.

```{r}
clss <- getClasses(object = lgb_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)

clss <- getClasses(object = lgb_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = binaryThresholdClass(threshold = 0.8,
                                                        positive_class = "positive"))
# By specifying such a high threshold for the positive class, predictions become "negative"
head(clss)
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = lgb_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = lgb_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "lgb_cls.mlu"
modelSave(object = lgb_cls_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### RangerClassifierModeler

Implementation of ranger in RMLutils for classification tasks. A fast Random Forest implementation for R.

#### Installation of ranger

First, we must check if the `ranger` library is installed:

```{r}
require("ranger")
```

If it is not, we can easily install it from CRAN,
using the command `install.packages("ranger")`.

#### Instantiation

The rangerClassifierModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `ranger`. It is recommended to consult the ranger documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

```{r}
# Define model parameters
target <- "classification_target"
regressors <- setdiff(colnames(c_data.dt), target)
params <- list(
  "num.trees" = 50,
  "max.depth" = 10
)
# Instantiate the model
ranger_cls_modeler <- rangerClassifierModeler(params = params,
                                              target = target,
                                              regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = ranger_cls_modeler,
              data = c_data.dt, # Complete data
              CVSampler = cvSampler(key = c_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

After determining a suitable set of parameters with crossValidate, we can train the chosen model.

```{r}
ranger_cls_modeler <- train(object = ranger_cls_modeler,
                            data = c_data.dt,
                            subSet = c_train.keys)
```

#### getPredictions

When predicting, classifier modelers return a matrix of scores:

```{r}
preds <- getPredictions(object = ranger_cls_modeler,
                        data = c_data.dt,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

A particular class mapping can be specified for this object, of which there are:

- `maxScoreClass()`: (default), chooses the class with the highest score.

- `binaryThresholdClass()`: Only for binary target. Allows specifying a selection threshold for the positive class.

```{r}
clss <- getClasses(object = ranger_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)

clss <- getClasses(object = ranger_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = binaryThresholdClass(threshold = 0.95,
                                                        positive_class = "positive"))
# By specifying such a high threshold for the positive class, some predictions become "negative"
head(clss)
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = ranger_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = ranger_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "ranger_cls.mlu"
modelSave(object = ranger_cls_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### catboostClassifierModeler

Implementation of catboost in RMLutils for classification tasks (open-source library for gradient boosting on decision trees, with built-in support for categorical variables).

#### Installation of catboost

First, we must check if the catboost library is installed:

```{r}
require("catboost")
```

Unlike other packages, catboost is not on the CRAN repository,
so to install it we must follow the instructions in this [link](https://catboost.ai/docs/en/concepts/r-installation). If it does not work,
it is recommended to download the compressed library (the tar.gz file), unzip it and run the following command:

```
install.packages("/path/to/unzipped/folder", 
repos = NULL, 
type = "source",
INSTALL_opts = c("--no-multiarch"))
```

Some recent versions may not work, in that case, we recommend downloading an older version and checking that the DESCRIPTION folder exists in the unzipped folder. Our testing was mostly done on version 1.2.8 of catboost for windows.

#### Instantiation

The catboostClassifierModeler requires the following parameters to be instantiated:

`params`: list with the names of the parameters required for `catboost.train`. It is recommended to consult the catboost documentation for more information.
In the case of classification, the `objective` parameter can be `"MultiClass"`/`"MultiClassOneVsAll"` (which works for both multiclass and binary target) or `"logloss"`/`"CrossEntropy"` (only for binary target).

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

- `val_data`: Validation dataset to use during training.

- `positive_class`: (Optional) For the case of objective = "logloss" or objective = "CrossEntropy", it is recommended to specify the name or identifier of the positive class, otherwise the order according to the factor levels will be assumed, with the first level being the negative class and the second the positive one.

- `ntree_start`: Additional parameter for catboost predict. 0 by default.

- `ntree_end`: Additional parameter for catboost predict. 0 by default.

- `thread_count`: Additional parameter for catboost predict. -1 by default.

```{r}
# Define model parameters
target <- "classification_target"
regressors <- setdiff(colnames(c_data.dt), target)
params <- list(
  # loss_function = 'MultiClass',
  # loss_function = 'MultiClassOneVsAll',
  loss_function = 'Logloss', # Binary
  # loss_function = 'CrossEntropy', # Binary
  iterations = 50,
  metric_period=10)
# Instantiate the model
catboost_cls_modeler <- catboostClassifierModeler(params = params,
                                                  target = target,
                                                  regressors = regressors,
                                                  positive_class = "positive")
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = catboost_cls_modeler,
              data = c_data.dt, # Complete data
              CVSampler = cvSampler(key = c_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

After determining a suitable set of parameters with crossValidate, we can train the chosen model.

```{r}
catboost_cls_modeler <- train(object = catboost_cls_modeler,
                              data = c_data.dt,
                              subSet = c_train.keys)
```

#### getPredictions

When predicting, classifier modelers return a matrix of scores:

```{r}
preds <- getPredictions(object = catboost_cls_modeler,
                        data = c_data.dt,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

A particular class mapping can be specified for this object, of which there are:

- `maxScoreClass()`: (default), chooses the class with the highest score.

- `binaryThresholdClass()`: Only for binary target. Allows specifying a selection threshold for the positive class.

```{r}
clss <- getClasses(object = catboost_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)

clss <- getClasses(object = catboost_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = binaryThresholdClass(threshold = 0.95,
                                                        positive_class = "positive"))
# By specifying a high threshold for the positive class, predicted classes remain the same, because the model predicts classes with high confidence (score close to 1)
head(clss)
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = catboost_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = catboost_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "catboost_cls.mlu"
modelSave(object = catboost_cls_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### miceClassifierModeler

Implementation of MICE (Multivariate Imputation by Chained Equations) in RMLutils for classification tasks. Although it is originally designed for missing data imputation, it can be adapted for prediction tasks.

**Important: mice output is stochastic. This means that even if trained and predicted on the same dataset, its output will generally be different. The user must take this into account when using it.**

#### Installation of mice

First, we must check if the `mice` library is installed:

```{r}
require("mice")
```

If it is not, we can easily install it from CRAN,
using the command `install.packages("mice")`.

#### Instantiation

The miceClassifierModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `mice`. It is recommended to consult the mice documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

```{r}
# Define model parameters
target <- "classification_target"
regressors <- setdiff(colnames(c_data.dt), target)
params <- list(m = 1,
               method = "rf",
               print = FALSE)
# Instantiate the model
mice_cls_modeler <- miceClassifierModeler(params = params,
                                          target = target,
                                          regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = mice_cls_modeler,
              data = c_data.dt, # Complete data
              CVSampler = cvSampler(key = c_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

After determining a suitable set of parameters with crossValidate, we can train the chosen model.

```{r}
mice_cls_modeler <- train(object = mice_cls_modeler,
                          data = c_data.dt,
                          subSet = c_train.keys)
```

#### getPredictions

When predicting, the mice classifier model returns a sparse matrix of ones and zeros, since it cannot return scores:

```{r}
preds <- getPredictions(object = mice_cls_modeler,
                        data = c_data.dt,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

```{r}
clss <- getClasses(object = mice_cls_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = mice_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = mice_cls_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "mice_cls.mlu"
modelSave(object = mice_cls_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

## Regression modelers

```{r}
# Define common data for the regression task
r_data.dt <- generateRegressionData(n_samples = 1000,
                                    generate_random_weights = FALSE)

# Define train/test partitions
set.seed(273)
r_data.keys <- 1:nrow(r_data.dt)
r_train_size = round(0.8 * nrow(r_data.dt))
r_smplr <- srsSampler(key = r_data.keys, ssize = r_train_size)
r_train.keys <- getSample(r_smplr)
r_test.keys <- setdiff(r_data.keys, r_train.keys)
head(r_data.dt)
```

### directRegressorModeler

A direct imputation modeler that scales a proto-target variable (assumed close to target) with a factor to be specified by the user. It does not require training.

```{r}
# Define a protoTarget on the data we have
drm_data.dt <- copy(r_data.dt)

# Calculate the median of the target variable according to the value of the "cat_3" variable in the training set
median_by_cat3.dt <- drm_data.dt[r_train.keys, .(proto_target = median(regression_target)), by = c("cat_3")]

# Merge with the original dataset
# The proto_target is the median according to the value of cat_3 on a sample of the data
drm_data.dt <- merge(drm_data.dt, median_by_cat3.dt, by = c("cat_3"))
```

####Instantiation

To instantiate a directRegressorModeler the following arguments are required:

- `protoTarget`: Name of the proto-target variable.

- `scaleFactor`: Scale factor to apply on protoTarget. It is set to 1 by default.

- `target`: Name of the target variable.

```{r}
# Assume a scaleFactor of 1, so our model simply returns protoTarget
d_reg_modeler <- directRegressorModeler(protoTarget = "proto_target",
                                        scaleFactor = 1,
                                        target = "regression_target")
```

#### getPredictions

Regressor modelers return a matrix with the predictions.

```{r}
preds <- getPredictions(object = d_reg_modeler,
                        data = drm_data.dt,
                        subSet = r_test.keys)
head(preds)
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = d_reg_modeler,
         data = drm_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = d_reg_modeler,
         data = drm_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "drm_reg.mlu"
modelSave(object = d_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = drm_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### HTRegressorModeler

A regression modeler that assumes the target variable is approximately proportional to its sampling weight.

```{r}
# Define appropriate data for this modeler
set.seed(273)
ht_data.dt <- generateStratifiedDomainEstimateData(n_samples = 100, generate_random_weights = FALSE)
# A variable is selected and sampling weights are defined proportional to its size
ht_data.dt <- ht_data.dt[, c("income"), with = FALSE]
# Sample size
n <- round(0.5 * nrow(ht_data.dt))
# Variables with pi >= 1 change their selection probability to 1
ht_data.dt[, pi := n*income/sum(income)]
ht_data.dt[, pi := ifelse(pi >= 1, 1, pi)]
# Definition of sampling weights
ht_data.dt[, weights := 1/pi]
head(ht_data.dt)
# A sample is taken proportional to its selection probabilities
ht_data.keys <- 1:nrow(ht_data.dt)
ht_train.keys <- sample(x = ht_data.keys, size = n, replace = FALSE, prob = ht_data.dt[, pi])
ht_test.keys <- setdiff(ht_data.keys, ht_train.keys)
```

#### Instantiation

To instantiate `HTRegressorModeler` the following arguments are required:

- `weightsColumn`: Name of the column with sampling weights.

- `target`: Name of the target variable.

```{r}
target <- "income"
weights_column <- "weights"

# Modeler definition
ht_reg_modeler <- HTRegressorModeler(target = target, weightsColumn = weights_column)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds. (Although its usefulness for this model is debatable).

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = ht_reg_modeler,
              data = ht_data.dt, # Complete data
              CVSampler = cvSampler(key = ht_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
ht_reg_modeler <- train(object = ht_reg_modeler,
                        data = ht_data.dt,
                        subSet = ht_train.keys)
```

#### getPredictions

Regressor modelers return a matrix with the predictions.

```{r}
preds <- getPredictions(object = ht_reg_modeler,
                        data = ht_data.dt,
                        subSet = ht_test.keys)
head(preds)
# Compare with actual values
matrix(head(ht_data.dt[ht_test.keys, income]))
```

#### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = ht_reg_modeler,
         data = ht_data.dt,
         cost_function = cf,
         subSet = ht_train.keys)
# Evaluation on the test set
evaluate(object = ht_reg_modeler,
         data = ht_data.dt,
         cost_function = cf,
         subSet = ht_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "ht_reg.mlu"
modelSave(object = ht_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = ht_data.dt,
         cost_function = cf,
         subSet = ht_test.keys)
```

### strataHTRegressorModeler

Version of HTRegressorModeler for cases with stratified data.

```{r}
# Define appropriate data for this modeler
set.seed(273)
str_ht_data.dt <- generateStratifiedDomainEstimateData(n_samples = 100, generate_random_weights = FALSE)
# A variable is selected and sampling weights are defined proportional to its size
str_ht_data.dt <- str_ht_data.dt[, c("income", "strata_class"), with = FALSE]
# Sample size
n <- round(0.5 * nrow(str_ht_data.dt))
# Variables with pi >= 1 change their selection probability to 1
str_ht_data.dt[, pi := n*income/sum(income)]
str_ht_data.dt[, pi := ifelse(pi >= 1, 1, pi)]
# Definition of sampling weights
str_ht_data.dt[, weights := 1/pi]
head(str_ht_data.dt)
# A sample is taken proportional to its selection probabilities
str_ht_data.keys <- 1:nrow(str_ht_data.dt)
str_ht_train.keys <- sample(x = str_ht_data.keys, size = n, replace = FALSE, prob = str_ht_data.dt[, pi])
str_ht_test.keys <- setdiff(str_ht_data.keys, str_ht_train.keys)
```

#### Instantiation

To instantiate strataHTRegressorModeler the following arguments are required:

- `weightsColumn`: Name of the column with sampling weights.

- `target`: Name of the target variable.

- `strata`: Name of the stratification column.

```{r}
target <- "income"
weights_column <- "weights"
strata_column <- "strata_class"

str_ht_reg_modeler <- strataHTRegressorModeler(weightsColumn = weights_column,
                                               target = target,
                                               strata = strata_column)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds. (Although its usefulness for this model is debatable, except if different columns are used for stratification).

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = str_ht_reg_modeler,
              data = str_ht_data.dt, # Complete data
              CVSampler = cvSampler(key = str_ht_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
str_ht_reg_modeler <- train(object = str_ht_reg_modeler,
                            data = str_ht_data.dt,
                            subSet = str_ht_train.keys)
```

#### getPredictions

Regressor modelers return a matrix with the predictions.

```{r}
preds <- getPredictions(object = str_ht_reg_modeler,
                        data = str_ht_data.dt,
                        subSet = str_ht_test.keys)
head(preds)
# Compare with actual values
matrix(head(str_ht_data.dt[str_ht_test.keys, income]))
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = str_ht_reg_modeler,
         data = str_ht_data.dt,
         cost_function = cf,
         subSet = str_ht_train.keys)
# Evaluation on the test set
evaluate(object = str_ht_reg_modeler,
         data = str_ht_data.dt,
         cost_function = cf,
         subSet = str_ht_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "str_ht_reg.mlu"
modelSave(object = str_ht_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = str_ht_data.dt,
         cost_function = cf,
         subSet = str_ht_test.keys)
```

### strataMeanRegressorModeler

A simple regression modeler that returns the average of the target according to several regressors used as stratification variables. The regressors must be categorical variables.

```{r}
# Define appropriate data for this modeler
smr_data.dt <- generateStratifiedDomainModelAssistedEstimateData(n_samples = 1000,
                                                             generate_random_weights = TRUE)
head(smr_data.dt)

# Define train and test partitions
smr_data.keys <- 1:nrow(smr_data.dt)
train_size = round(0.8 * nrow(smr_data.dt))
smplr <- srsSampler(key = smr_data.keys, ssize = train_size)
smr_train.keys <- getSample(smplr)
smr_test.keys <- setdiff(smr_data.keys, smr_train.keys)
```

#### Instantiation

The strataMeanRegressorModeler requires the following arguments to be instantiated:

- `target`: Name of the target variable.

- `regressors`: Vector with the names of the stratification variables.

- `weightsColumn`: Name of the sampling weights column.

```{r}
target <- "income" # Regression
weightsColumn <- "weights"
regressors <- c("strata_class", "domain_class") # Only categorical variables

# Instantiate the model
sm_reg_modeler <- strataMeanRegressorModeler(target = target,
                                             regressors = regressors,
                                             weightsColumn = "weights")
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds. (Although its usefulness for this model is debatable, except if different columns are used for stratification).

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = sm_reg_modeler,
              data = smr_data.dt, # Complete data
              CVSampler = cvSampler(key = smr_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
sm_reg_modeler <- train(object = sm_reg_modeler,
                            data = smr_data.dt,
                            subSet = smr_train.keys)
```

#### getPredictions

Regressor modelers return a matrix with the predictions.

```{r}
preds <- getPredictions(object = sm_reg_modeler,
                        data = smr_data.dt,
                        subSet = smr_test.keys)
head(preds)
# Compare with actual values
matrix(head(smr_data.dt[smr_test.keys, income]))
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = sm_reg_modeler,
         data = smr_data.dt,
         cost_function = cf,
         subSet = smr_train.keys)
# Evaluation on the test set
evaluate(object = sm_reg_modeler,
         data = smr_data.dt,
         cost_function = cf,
         subSet = smr_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "sm_reg.mlu"
modelSave(object = sm_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = smr_data.dt,
         cost_function = cf,
         subSet = smr_test.keys)
```


### LGBRegressorModeler

Implementation of LightGBM in RMLutils for classification tasks (lightweight gradient boosting framework with tree-based models).

#### Installation of lightgbm

First, we must check if the lightgbm library is installed:

```{r}
require("lightgbm")
```

If it is not, we can easily install it from CRAN, using the command `install.packages("lightgbm")`.

#### Instantiation

The LGBRegressorModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `lgb.train.` It is recommended to consult the lightgbm documentation for more information.
In the case of regression, the `objective` parameter must be `"regression"`.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the regression task.

```{r}
# Define model parameters
target <- "regression_target"
regressors <- setdiff(colnames(r_data.dt), target)
params <- list(
  "learning_rate" = 0.07,
  "nrounds" = 10,
  "objective" = "regression"
)
lgb_reg_modeler <- LGBRegressorModeler(params = params,
                                       target = target,
                                       regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = lgb_reg_modeler,
              data = r_data.dt, # Complete data
              CVSampler = cvSampler(key = r_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
lgb_reg_modeler <- train(object = lgb_reg_modeler,
                         data = r_data.dt,
                         subSet = r_train.keys)
```

#### getPredictions

Regressor modelers return a matrix with the predictions.

```{r}
preds <- getPredictions(object = lgb_reg_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = lgb_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = lgb_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "lgb_reg.mlu"
modelSave(object = lgb_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### RangerRegressorModeler

#### Installation of ranger

First, we must check if the `ranger` library is installed:

```{r}
require("ranger")
```

If it is not, we can easily install it from CRAN,
using the command `install.packages("ranger")`.

#### Instantiation

The `rangerRegressorModeler` requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `ranger`. It is recommended to consult the ranger documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the regression task.

```{r}
# Define model parameters
target <- "regression_target"
regressors <- setdiff(colnames(r_data.dt), target)
params <- list(
  "num.trees" = 50,
  "max.depth" = 10
)
# Instantiate the model
ranger_reg_modeler <- rangerRegressorModeler(params = params,
                                             target = target,
                                             regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = ranger_reg_modeler,
              data = r_data.dt, # Complete data
              CVSampler = cvSampler(key = r_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
ranger_reg_modeler <- train(object = ranger_reg_modeler,
                            data = r_data.dt,
                            subSet = r_train.keys)
```

#### getPredictions

```{r}
preds <- getPredictions(object = ranger_reg_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = ranger_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = ranger_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "ranger_reg.mlu"
modelSave(object = ranger_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### catboostRegressorModeler

Implementation of catboost in RMLutils for regression tasks (open-source library for gradient boosting on decision trees, with built-in support for categorical variables).

#### Installation of catboost

First, we must check if the catboost library is installed:

```{r}
require("catboost")
```

Unlike other packages, catboost is not on the CRAN repository,
so to install it we must follow the instructions in this [link](https://catboost.ai/docs/en/concepts/r-installation). If it does not work,
it is recommended to download the compressed library (the tar.gz file), unzip it and run the following command:

```
install.packages("/path/to/unzipped/folder", 
repos = NULL, 
type = "source",
INSTALL_opts = c("--no-multiarch"))
```

Some recent versions may not work, in that case, we recommend downloading an older version and checking that the DESCRIPTION folder exists in the unzipped folder. Our testing was mostly done on version 1.2.8 of catboost for windows.

#### Instantiation

The catboostRegressorModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `catboost.train`. It is recommended to consult the catboost documentation for more information.
In the case of regression, the `objective` parameter can take multiple values, such as `"RMSE"`, `"MAE"`, `"MAPE"`, etc.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the regression task.

- `val_data`: Validation dataset to use during training.

- `ntree_start`: Additional parameter for catboost predict. 0 by default.

- `ntree_end`: Additional parameter for catboost predict. 0 by default.

- `thread_count`: Additional parameter for catboost predict. -1 by default.

```{r}
# Define model parameters
target <- "regression_target"
regressors <- setdiff(colnames(r_data.dt), target)
params <- list(
  loss_function = 'RMSE',
  iterations = 50,
  metric_period=10)
# Instantiate the model
catboost_reg_modeler <- catboostRegressorModeler(params = params,
                                                 target = target,
                                                 regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = catboost_reg_modeler,
              data = r_data.dt, # Complete data
              CVSampler = cvSampler(key = r_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
catboost_reg_modeler <- train(object = catboost_reg_modeler,
                              data = r_data.dt,
                              subSet = r_train.keys)
```

#### getPredictions

```{r}
preds <- getPredictions(object = catboost_reg_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = catboost_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = catboost_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "catboost_reg.mlu"
modelSave(object = catboost_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### miceRegressorModeler

Implementation of MICE (Multivariate Imputation by Chained Equations) in RMLutils for regression tasks. Although it is originally designed for missing data imputation, it can be adapted for prediction tasks.

**Important: mice output is stochastic. This means that even if trained and predicted on the same dataset, its output will generally be different. The user must take this into account when using it.**

#### Installation of mice

First, we must check if the `mice` library is installed:

```{r}
require("mice")
```

If it is not, we can easily install it from CRAN,
using the command `install.packages("mice")`.

#### Instantiation

The miceRegressorModeler requires the following parameters to be instantiated:

- `params`: list with the names of the parameters required for `mice`. It is recommended to consult the mice documentation for more information.

- `target`: Name of the target variable in the dataset.

- `regressors`: Vector of strings indicating the names of the columns to use as variables for the classification task.

```{r}
# Define model parameters
target <- "regression_target"
regressors <- setdiff(colnames(r_data.dt), target)
params <- list(m = 1,
               method = "rf",
               print = FALSE)
# Instantiate the model
mice_reg_modeler <- miceRegressorModeler(params = params,
                                          target = target,
                                          regressors = regressors)
```

#### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- MSECostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = mice_reg_modeler,
              data = r_data.dt, # Complete data
              CVSampler = cvSampler(key = r_train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

#### train

```{r}
mice_reg_modeler <- train(object = mice_reg_modeler,
                          data = r_data.dt,
                          subSet = r_train.keys)
```

#### getPredictions

```{r}
preds <- getPredictions(object = mice_reg_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

#### evaluate

```{r}
# Define a cost function
cf <- MSECostFunction()
# Evaluation on the training set
evaluate(object = mice_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = mice_reg_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

#### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "mice_reg.mlu"
modelSave(object = catboost_reg_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

## Using expandHyperParams

This is a utility function to generate sets of hyperparameters for our models. It is particularly suitable when we want to apply cross-validation on multiple models with different hyperparameters.

Next, we will apply an optimal hyperparameter search process for a regression problem with `rangerRegressorModeler`:

```{r}
# Define the data
data.dt <- generateStratifiedDomainModelAssistedEstimateData(n_samples = 5000,
                                                             generate_random_weights = FALSE)

# Define train/test partitions
set.seed(273)
data.keys <- 1:nrow(data.dt)
train_size = round(0.8 * nrow(data.dt))
smplr <- srsSampler(key = data.keys, ssize = train_size)
train.keys <- getSample(smplr)
test.keys <- setdiff(data.keys, train.keys)
head(data.dt)

# Define set of hyperparameters we want to test
target <- "target"
regressors <- setdiff(colnames(data.dt), target)
hyperparams <- list(
  "num.trees" = c(10, 25, 50),
  "max.depth" = c(5, 10),
  "replace" = c(TRUE, FALSE),
  "alpha" = seq(0,1,0.1)
)

# Call the expandHyperParams function to generate all combinations
combinations <- expandHyperParams(hyperparams)
# Total combinations:
print(paste0("Total combinations: ", length(combinations)))

# Choose the hyperparameters that minimize the MSE in training by applying cross-validation
cf <- MSECostFunction()
n_folds <- 5
min_mse_idx <- NA
min_mse <- Inf
for (i in 1:length(combinations)){

  # Define modeler with the current hyperparameter combination
  modeler <- rangerRegressorModeler(params = combinations[[i]],
                                    target = target,
                                    regressors = regressors)
  # Apply cross-validation on the training set
  cost <- crossValidate(modeler = modeler,
                        data = data.dt, # Complete data
                        CVSampler = cvSampler(key = train.keys, folds = n_folds),
                        costFunction = cf)
   if(cost$target < min_mse){
     min_mse_idx <- i
     min_mse <- cost$target
   }
}
print("Best hyperparameters: ")
print(combinations[[min_mse_idx]])
print("Best MSE:")
print(min_mse)

# Train rangerRegressorModeler with those hyperparameters and
# the full training set
modeler <- rangerRegressorModeler(params = combinations[[min_mse_idx]],
                                  target = target,
                                  regressors = regressors)
modeler <- train(object = modeler,
                 data = data.dt,
                 subSet = train.keys)

# Evaluate on the training set and on test
# Evaluation on the training set
evaluate(object = modeler,
         data = data.dt,
         cost_function = cf,
         subSet = train.keys)
# Evaluation on the test set
evaluate(object = modeler,
         data = data.dt,
         cost_function = cf,
         subSet = test.keys)

# Compare with the R^2 cost function
cf2 <- r2CostFunction()
# Evaluation on the training set
evaluate(object = modeler,
         data = data.dt,
         cost_function = cf2,
         subSet = train.keys)
# Evaluation on the test set
evaluate(object = modeler,
         data = data.dt,
         cost_function = cf2,
         subSet = test.keys)
```

We can see that the model fits the data reasonably well, and that most of the variance of the data is explained by the modeling.
