---
title: "prePredModeler tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler and Modeler tutorials before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

## The prePredRegressorModeler object

This object is one of the complex modelers in RMLutils. It is a two-step model (also known as a chain model), which internally uses two regression modelers, called predModeler and mainModeler.

The predModeler is responsible for predicting a target variable (or multiple ones) which the mainModeler then incorporates into its regressors for the main prediction.

```{r}
# Generate a dataset
data.dt <- generateRegressionData(n_samples = 1000, generate_random_weights = FALSE)
# Create an additional column that depends on regression_target, which will be the target variable of the modeler
data.dt[, main_regression_target := regression_target + float_1 - float_2]

# Define train and test partitions
set.seed(273)
data.keys <- 1:nrow(data.dt)
train_size = round(0.8 * nrow(data.dt))
smplr <- srsSampler(key = data.keys, ssize = train_size)
train.keys <- getSample(smplr)
test.keys <- setdiff(data.keys, train.keys)
```

### Instantiation

The prePredRegressorModeler requires the following parameters to be instantiated:

- `predModeler`: A univariate or multivariate modeler to generate the additional regressor(s) that mainModeler will use for its prediction task.

- `mainModeler`: The main modeler. It will use the predictions from predModeler together with other regressors to predict the target variable.

For this tutorial, ranger modelers will be used to define the prePredModeler. The user should check that the `ranger` library is installed.

We define the previous model and the main one:

```{r}
# Define the previous modeler (this will have regression_target as target variable)
params <- list(
  "num.trees" = 50,
  "max.depth" = 20
)
target <- "regression_target"
regressors <- setdiff(colnames(data.dt), c(target, "main_regression_target"))

predModeler <- rangerRegressorModeler(params = params,
                                      target = target,
                                      regressors = regressors)

# Define the main modeler
# This will predict main_regression_target, taking regression_target as an additional regressor
params <- list(
  "num.trees" = 50,
  "max.depth" = 20
)
target <- "main_regression_target"
regressors <- setdiff(colnames(data.dt), c(target))

mainModeler <- rangerRegressorModeler(params = params,
                                      target = target,
                                      regressors = regressors)
```

Now we define the prePredModeler:

```{r}
prePredModeler <- prePredRegressorModeler(predModeler = predModeler,
                                          mainModeler = mainModeler)
```


### crossValidate

```{r}
# Define a cost function
cf <- r2CostFunction()
# Apply cross-validation on the training set
crossValidate(modeler = prePredModeler,
              data = data.dt, # Complete data
              CVSampler = cvSampler(key = train.keys, folds = 3), # cvSampler on training keys
              costFunction = cf)
```

### train

```{r}
prePredModeler <- train(prePredModeler,
                        data = data.dt,
                        subSet = train.keys)
```

### getPredictions

```{r}
preds <- getPredictions(object = prePredModeler,
                        data = data.dt,
                        subSet = test.keys)
head(preds)
```

### evaluate

Since this modeler contains two other modelers, we may want to evaluate the final output, or evaluate both modelers at the same time. The evaluate method allows us to specify a regression cost function (which will evaluate only the final output), or a named list with cost functions to specify the cost function with which to evaluate each of the internal modelers:

```{r}
# Define a cost function
cf <- r2CostFunction()
# Evaluation on the training set
evaluate(object = prePredModeler,
         data = data.dt,
         cost_function = cf,
         subSet = train.keys)
# Evaluation on the test set
evaluate(object = prePredModeler,
         data = data.dt,
         cost_function = cf,
         subSet = test.keys)


# Evaluation of both modelers separately
cf2 <- list("predModeler" = r2CostFunction(),
            "mainModeler" = r2CostFunction())


# Evaluation on the training set for both modelers
evaluate(object = prePredModeler,
         data = data.dt,
         cost_function = cf2,
         subSet = train.keys)
# Evaluation on the test set for both modelers
evaluate(object = prePredModeler,
         data = data.dt,
         cost_function = cf2,
         subSet = test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "pre_pred_reg.mlu"
modelSave(object = prePredModeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = data.dt,
         cost_function = cf,
         subSet = test.keys)
```

