---
title: "preproModeler tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
dir.create(file.path("./trained_modelers"), showWarnings = FALSE)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

It is recommended that the user reviews the Sampler and Modeler tutorials before reading this document.

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

**Additionally, it is important to emphasize that, to use most modelers, the user will be responsible for installing the specific library they use. To keep a lightweight installation, RMLutils does not install any of the machine learning libraries used by the modelers.**

## The preproModeler, preproObject and preproFunction objects

PreproModelers are one of the types of complex modelers in RMLutils, which work for both classification and regression tasks.

The main characteristic of PreproModelers is that they apply a preprocessing function just before training or predicting with any modeler.
This preprocessing function can be fully defined by the user to suit their needs.
In order to integrate such a function with the RMLutils modeler interface, the user must make use of the preproObject. In the following code segment, one is defined with a preprocessing function already implemented in RMLutils:

```{r}
# Definition of preprocessing parameters
prepro_params <- list(modeler_type = NULL, pretrained_min_max = NULL)

# Definition of preproObject
prepro_object <- preproObject(preproFunction = minMaxScalerPreproFunction,
                              params = prepro_params)
```

The `minMaxScalerPreproFunction` scales the numeric variables of the dataset so that they are in the 0-1 range. Once the minimums and maximums have been calculated internally (necessary for scaling), they are saved in the preproObject parameters.

## Format of the preprocessing function

The preprocessing function must comply with several characteristics:

1) It must take a preproObject as its only input. All parameters that the preprocessing function may need must go inside the prepro_params attribute of the PreproObject.

2) All necessary parameters within prepro_params can be altered (even deleted or new ones added), but altering any other attribute of PreproObject should be avoided, except for the exception in the following point.

3) The preprocessing function must overwrite the new_regressors attribute in preproObject, so that the modelers know which regressors to use. If the preprocessing function does not create/delete columns, the old regressors should be assigned as new_regressors. Likewise, the data attribute will also be altered.

4) The output of the preprocessing function will be the preproObject itself, with its data attribute overwritten with the preprocessed data and new_regressors specified. Its params attribute may also be altered.

If we want to preprocess the data externally (without using preproModelers), we must proceed as follows:

```{r}
# We do not instantiate the prepro_object again, we can use the previous one

# Define the data
data.dt <- generateRegressionData(n_samples = 1000,
                                  generate_random_weights = TRUE)
head(data.dt) # We have sampling weights. We do not want to alter these

# We are going to define a "fake" modeler, as a list with the target, regressors and weightsColumn attributes
target <- "regression_target"
weightsColumn <- "weights"
regressors <- setdiff(colnames(data.dt), c(target, weightsColumn))
modeler <- list("target" = target,
                "regressors" = regressors,
                "weightsColumn" = weightsColumn)

# Assign these data to the prepro_object, using the assignExternalAttributes function
prepro_object <- assignExternalAttributes(modeler = modeler,
                                          object = prepro_object)

# Call getPreproData to return the preproObject with the preprocessed data
prepro_object <- getPreproData(object = prepro_object, data = data.dt)

# Since we specified the weights column, that one has not been altered (the same happens with target)
head(prepro_object$data)
```

Below is the code of the preprocessing function we have used, so that the user can get an idea of how to program their own.

```
minMaxScalerPreproFunction <- function(object = NULL){

  # Extract the necessary variables from the preproObject
  data <- object$data
  target <- object$target
  regressors <- object$regressors
  weightsColumn <- object$weightsColumn
  modeler_type <- object$params$modeler_type
  pretrained_min_max <- object$params$pretrained_min_max
  
  # modeler_type is there for compatibility with the rest of RMLutils modelers.
  # With this we will assign a new_regressors attribute correctly
  # according to the modeler type (at the end of the function).
  if(is.null(modeler_type)){
    modeler_type <- "simple"
  }

  dt <- as.data.table(data)

  # Scan the dataset to detect the names of numeric columns
  num_vars <- colnames(dt)[sapply(dt, is.numeric)]

  # Define which columns should not be preprocessed (in our case, target and weightsColumn should not be touched)
  if(any(target %in% colnames(data))){
    if(!is.null(weightsColumn)){
      if(weightsColumn %in% colnames(data)){
        no_preprocessing <- c(target, weightsColumn)
      }else{
        no_preprocessing <- c(target)
      }
    }else{
      no_preprocessing <- c(target)
    }

  }else{
    if(!is.null(weightsColumn)){
      if(weightsColumn %in% colnames(data)){
        no_preprocessing <- c(weightsColumn)
      }else{
        no_preprocessing <- c()
      }
    }else{
      no_preprocessing <- c()
    }
  }

  # Remove the numeric variables that should not be preprocessed (they are in no_preprocessing)
  # (i.e. weightsColumn and target)
  num_vars <- num_vars[!num_vars %in% no_preprocessing]

  # The minimum and maximum are calculated for each numeric variable
  # This will be saved in the preproObject so that in later executions it scales
  # the data with the first minimums and maximums it detects (and does not have to calculate them each time)
  if(is.null(pretrained_min_max)){
    pretrained_min_max <- lapply(num_vars, function(var){
      return(list("min" = min(dt[[var]]),
                  "max" = max(dt[[var]])))
    })
    names(pretrained_min_max) <- num_vars

    # Save in the preproObject parameters
    object$params$pretrained_min_max <- pretrained_min_max
  }else{
    # If there already are, we just assign them, we do not generate them again
    pretrained_min_max <- object$params$pretrained_min_max
  }

  # Scale the numeric variables to the 0-1 range (MinMax scaling)
  # The scaled variables are renamed to "st_" + "<original_name>"
  for (var in num_vars){
    dt[, paste0("st_", var) := (get(var) - pretrained_min_max[[var]]$min) / (pretrained_min_max[[var]]$max - pretrained_min_max[[var]]$min)]
    dt[, (var) := NULL]
  }

  # Overwrite the data attribute with the preprocessed data
  object$data <- dt

  # Define the name of the new regressors and add them to the preproObject
  prepro_regressors <- colnames(dt)
  prepro_regressors<- prepro_regressors[! prepro_regressors %in% no_preprocessing]
  switch (modeler_type,
    "simple" = { # For all other modelers that do not have more than one internal modeler or are a simple model
      object$new_regressors <- prepro_regressors
      },
    "classifRegModeler" = {
      object$new_regressors <- list(
                                  "regModeler" = prepro_regressors,
                                  "classifModeler" = prepro_regressors)
    },
    "prePredModeler" = {
      object$new_regressors <- list(
                                  "predModeler" = prepro_regressors,
                                  # Add target of predModeler to regressors
                                  "mainModeler" = c(target[1], prepro_regressors))
    },
    "benchModeler" = {
      object$new_regressors <- lapply(1:length(regressors), function(x){
                                  return(prepro_regressors)
                                })
    },
    {stop(paste0("The provided modeler_type parameter: ", modeler_type, " was not recognized. Available types are \n NULL, classifRegModeler, prePredModeler and benchRegressorModeler"))}
  )

  return(object)
}
```

## preproClassifierModeler

```{r}
# Generate a dataset
c_data.dt <- generateMulticlassClassificationData(n_samples = 1000,
                                                  generate_random_weights = FALSE)

# Define train and test partitions
set.seed(273)
c_data.keys <- 1:nrow(c_data.dt)
c_train_size = round(0.8 * nrow(c_data.dt))
c_smplr <- srsSampler(key = c_data.keys, ssize = c_train_size)
c_train.keys <- getSample(c_smplr)
c_test.keys <- setdiff(c_data.keys, c_train.keys)
```

### Instantiation

The preproClassifierModeler requires the following parameters to be instantiated:

- `modeler`: Classification modeler to predict with the data after preprocessing. It must be instantiated like any other modeler.

- `preproObject`: preproObject object, containing a preprocessing function and parameters, which will store necessary attributes during the execution of the preprocessing function.

In this tutorial, a ranger modeler will be used to define the `preproClassifierModeler`. The user should check that the `ranger` library is installed.

We first define the internal modeler and the preproObject:

```{r}
c_target <- "classification_target"
c_regressors <- setdiff(colnames(c_data.dt), c_target)
c_params <- list(
  "num.trees" = 50,
  "max.depth" = 20
)

# Internal classification model
c_modeler <- rangerClassifierModeler(params = c_params,
                                     target = c_target,
                                     regressors = c_regressors)

# Preprocessing parameters
c_prepro_params <- list(modeler_type = NULL, pretrained_min_max = NULL)

# Define the preproObject
c_prepro_object <- preproObject(preproFunction = minMaxScalerPreproFunction,
                                params = c_prepro_params)
```

Finally, we define the preproModeler:

```{r}
pc_modeler <- preproClassifierModeler(modeler = c_modeler,
                                      preproObject = c_prepro_object)
```


### crossValidate

`crossValidate` returns the averaged cost for all folds.

```{r}
# Define a cost function
cf <- accuracyCostFunction()
crossValidate(modeler = pc_modeler,
              data = c_data.dt,
              CVSampler = cvSampler(key = c_train.keys, folds = 3),
              costFunction = cf)
```

### Train

```{r}
pc_modeler <- train(pc_modeler,
                    data = c_data.dt,
                    subSet = c_train.keys)
```

### getPredictions

When predicting, classifier modelers return a matrix of scores:

```{r}
preds <- getPredictions(object = pc_modeler,
                        data = c_data.dt,
                        subSet = c_test.keys)
head(preds)
```

To get the classes directly, we must use the `getClasses` method.

A particular class mapping can be specified for this object, of which there are:

`maxScoreClass()`: (default), chooses the class with the highest score.

`binaryThresholdClass()`: Only for binary target. Allows specifying a selection threshold for the positive class.

However, since it is a multiclass modeler, it only accepts the maxScoreClass() method.

```{r}
clss <- getClasses(object = pc_modeler,
                   data = c_data.dt,
                   subSet = c_test.keys,
                   class_mapping = maxScoreClass())
head(clss)
```

### evaluate

This function is a common interface for all modelers when evaluating their performance for a cost function and a specific dataset:

```{r}
# Define a cost function
cf <- accuracyCostFunction()
# Evaluation on the training set
evaluate(object = pc_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_train.keys)
# Evaluation on the test set
evaluate(object = pc_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "prepro_cls.mlu"
modelSave(object = pc_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = c_data.dt,
         cost_function = cf,
         subSet = c_test.keys)
```

## preproRegressorModeler

```{r}
# Generate a dataset
r_data.dt <- generateRegressionData(n_samples = 1000,
                                    generate_random_weights = FALSE)

# Define train and test partitions
set.seed(273)
r_data.keys <- 1:nrow(r_data.dt)
r_train_size = round(0.8 * nrow(r_data.dt))
r_smplr <- srsSampler(key = r_data.keys, ssize = r_train_size)
r_train.keys <- getSample(r_smplr)
r_test.keys <- setdiff(r_data.keys, r_train.keys)
```

### Instantiation

The preproRegressorModeler requires the following parameters to be instantiated:

- `modeler`: Regression modeler to predict with the data after preprocessing. It must be instantiated like any other modeler.

- `preproObject`: preproObject object, containing a preprocessing function and parameters, which will store necessary attributes during the execution of the preprocessing function.

In this tutorial, a ranger modeler will be used to define the preproRegressorModeler. The user should check that the `ranger` library is installed.

We first define the internal modeler and the preproObject:

```{r}
r_target <- "regression_target"
r_regressors <- setdiff(colnames(r_data.dt), r_target)

# Define a regressor modeler
r_params <- list(
  "num.trees" = 50,
  "max.depth" = 20
)
# Internal regression model
r_modeler <- rangerRegressorModeler(params = r_params,
                                    target = r_target,
                                    regressors = r_regressors)

# Preprocessing parameters
r_prepro_params <- list(modeler_type = NULL, pretrained_min_max = NULL)

# Define the preproObject
r_prepro_object <- preproObject(preproFunction = minMaxScalerPreproFunction,
                                params = r_prepro_params)
```

Finally, we define the preproModeler:

```{r}
pr_modeler <- preproRegressorModeler(modeler = r_modeler,
                                     preproObject = r_prepro_object)
```

### crossValidate

```{r}
# Define a cost function
cf <- r2CostFunction()
crossValidate(modeler = pr_modeler,
              data = r_data.dt,
              CVSampler = cvSampler(key = r_train.keys, folds = 3),
              costFunction = cf)
```

### Train

```{r}
pr_modeler <- train(pr_modeler,
                    data = r_data.dt,
                    subSet = r_train.keys)
```

### getPredictions

```{r}
preds <- getPredictions(object = pr_modeler,
                        data = r_data.dt,
                        subSet = r_test.keys)
head(preds)
```

### evaluate

```{r}
# Define a cost function
cf <- r2CostFunction()
# Evaluation on the training set
evaluate(object = pr_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_train.keys)
# Evaluation on the test set
evaluate(object = pr_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

### modelSave/modelLoad

```{r}
# Define a save path and a file name
m_path <- "./trained_modelers"
m_filename <- "prepro_reg.mlu"
modelSave(object = pr_modeler,
          path = m_path,
          fileName = m_filename)

# Load the model again and check that it works
loaded_modeler <- modelLoad(path = m_path, fileName = m_filename)
evaluate(object = loaded_modeler,
         data = r_data.dt,
         cost_function = cf,
         subSet = r_test.keys)
```

