---
title: "Sampler objects tutorial"
author: "Jordi Verdú Naranjo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: united
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RMLutils)
library(data.table)
```
<!-- 
# -*- coding: utf-8 -*-
#------------------------------------------------------------------------------#
# Copyright (C) [2026] Instituto Nacional de Estadística
#
# Este archivo forma parte del proyecto MLutils.
#
# Licenciado bajo la Licencia Pública de la Unión Europea (EUPL) v.1.2.
# Puede obtener una copia de la licencia en la raiz de este proyecto o en:
# https://eupl.eu/1.2/es/
#
# A menos que se indique lo contrario, este software se distribuye
# "TAL CUAL", SIN GARANTÍAS NI CONDICIONES DE NINGÚN TIPO.
# Consulte la licencia para conocer los términos específicos.
#------------------------------------------------------------------------------#
# Copyright (C) [2026] National Institute of Statistics
#
# This file is part of the MLutils project.
#
# Licensed under the European Union Public License (EUPL) v.1.2.
# You can obtain a copy of the license at the root of this project or at:
# https://eupl.eu/1.2/es/
#
# Unless otherwise indicated, this software is distributed
# "AS IS", WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND.
# See the license for specific terms.
#------------------------------------------------------------------------------#
-->

The user is reminded that all RMLutils methods and functions are
documented. Using the help tab, or the command itself, essential documentation
can be obtained for any object or function.

## Structure, types and methods of samplers

Samplers in RMLutils are objects that allow defining a sampling design, with
custom characteristics according to the dataset we wish to use.

Although some samplers require multiple arguments to instantiate,
all of them have a common argument, the `key` of the dataset.

Sampler objects have only one associated method: `get_sample()`,
which returns a sample (or multiple samples) according to the sampler
specifications.

Currently the following types exist:


* `srsSampler`
* `ssSampler`
* `bootstrapSampler`
* `fullSampler`
* `cvSampler`
* `cvGroupSampler`
* `cvMixedSampler`
* `cvTemporalSampler`

Below we generate a synthetic dataset, using one of the RMLutils functions and define its key:

```{r}
# Generate data with 20 rows
data.dt <- generateBinaryClassificationData(n_samples = 20,
                                            generate_random_weights = FALSE)

# Define the key as the vector of row indices of the dataset
key <- 1:nrow(data.dt)
print(key)
```

Now we can define a simple random sampling without replacement and
obtain samples:

```{r}
set.seed(273)
# ssize is the sample size
sampler <- srsSampler(key = key, ssize = 5)

# Obtain two independent samples by setting times = 2
samples <- getSample(sampler, times = 2)
print(samples)
```

Samples are returned as a 2D array, so if we want to keep only one,
we must select one column of the array.

```{r}
# Apply one of the samples to the dataset:
data.dt[samples[, 1]]
```

## srsSampler

This object allows defining a simple random sampling without replacement,
specifying the number of samples to obtain with its ssize parameter:

```{r}
srs_sampler <- srsSampler(key = 1:20,
                        ssize = 10)

set.seed(273)
# If we do not specify the times parameter, it is assumed to be 1.
getSample(srs_sampler)
```

If we wanted to obtain the complementary set (which could be a
validation/test set if we are training models), we can use
the base R `setdiff` function:

```{r}
set.seed(273)
train.keys <- getSample(srs_sampler)
test.keys <- setdiff(srs_sampler$key, train.keys)
print(train.keys)
print(test.keys)
```

## ssSampler

This object is the implementation of stratified random sampling (without
replacement) in RMLutils. It requires an additional argument to be defined,
a vector or array of strata that matches the dimensions of the dataset
we are going to use.

Additionally, the `ssize` argument must be a vector specifying the number
of samples to obtain from each stratum. See an example:

```{r}
# Define data with a variable that allows us to stratify
s_data.dt <- generateStratifiedDomainEstimateData(n_samples = 10)
print(s_data.dt)

# Then, define the sampler
ss_sampler <- ssSampler(key = 1:nrow(s_data.dt),
                        strata = s_data.dt[, strata_class],
                        ssize = c("Urban" = 3, "Rural" = 2))

# And obtain a sample
ss_sample <- getSample(ss_sampler)
print(s_data.dt[ss_sample])
```

## bootstrapSampler

This sampler generates bootstrap samples on a specified set of keys.
Its only parameter is `key`, since the sample size is fixed, as it matches
the size of `key`.

```{r}
b_sampler <- bootstrapSampler(key = 1:10)
set.seed(273)
getSample(b_sampler)
```

## fullSampler

A toy sampler that has two possible outputs. If its `ssize` parameter is
equal to `2`, it returns all possible combinations of pairs of elements from
the dataset. If `ssize` takes another value (or is left at the default),
it returns the full sample.

```{r}
key <- 1:6

# Return all possible pair combinations
f_smplr1 <- fullSampler(key = key, ssize = 2)

s1 <- getSample(f_smplr1)
print(s1)

# Return the full sample
f_smplr2 <- fullSampler(key = key)

s2 <- getSample(f_smplr2)
print(s2)
```


## cvSampler

The cvSampler generates samples without replacement for k-fold cross-validation,
so that each of its samples contains $k-1$ of the folds, and leaves one for
validation.

In case the dataset keys are not an integer multiple of the `folds` parameter,
some partitions will contain `NA`.

```{r}
# Define a cvSampler with 5 folds
smplr <- cvSampler(key = 1:20, folds = 5)
set.seed(273)
s1 <- getSample(smplr)

# Folds come out of size 4, so samples will contain 4x4 elements
print(s1)
```

Usually, this sampler will be used together with the `crossValidate` method of
modelers to perform cross-validation.

## cvGroupSampler

A variant of cvSampler that instead of partitioning the data randomly
according to their indices, generates its partitions according to a grouping or
stratification variable. That is, it will sample over the groups that exist in
the data, not over row indices.

```{r}
# Generate a dataset to visualize more clearly
set.seed(273)
# Define data (10 groups, 10 rows for each group)
# There are 10 groups, with 10 rows each
grouping_var <- c(sapply(paste0("Group_", 1:10), function(x){
return(rep(x, 10))
}, USE.NAMES = FALSE))
measure <- rnorm(n = length(grouping_var), mean = 10, sd = 1)
group_data.dt <- data.table(measure = measure, grouping_var = grouping_var)

head(group_data.dt)

# Define the sampler
key <- 1:nrow(group_data.dt)
identifier <- "grouping_var"
folds <- 3
cv_group_sampler <- cvGroupSampler(key = key,
                                   data = group_data.dt,
                                   identifier = identifier,
                                   folds = folds)
# As can be seen, the partitions contain all rows of the groups
# that they select
s1 <- getSample(cv_group_sampler)
print(s1)
table(group_data.dt[s1[, 1], grouping_var])
table(group_data.dt[s1[, 2], grouping_var])
table(group_data.dt[s1[, 3], grouping_var])
```


## cvMixedSampler

Another variant of cvSampler for cases where we want to force a fixed set
of rows to always be included in the samples, together with a variable set.

Apart from `key` and `folds`, it has an additional parameter called `fixedKey`,
which is the vector of fixed indices to always include in the sample. `key` must
include both the fixed and the variable keys.

```{r}
all_keys <- 1:50 # There are 50 elements in the sample
fixed_keys <- 1:20 # We want the first 20 to always be included

cv_mixed_sampler <- cvMixedSampler(key = all_keys,
                                   fixedKey = fixed_keys,
                                   folds = 5)
set.seed(273)
# As can be seen, the first 20 are always there
print(getSample(cv_mixed_sampler))
```

## cvTemporalSampler

An additional variant of cvSampler, for data with temporal reference or
periods of some kind, and where the output is required to be sorted
chronologically.

The samples it returns are an expanding and ordered temporal window. It is
useful for training models with a number of previous periods and checking
their performance on the most recent periods.

Its definition is more complex than the rest of the samplers, because it requires
the following arguments:

key: Row indices.

* `data`: Full dataset (as a data.table).
* `identifier`: String indicating the name of the column containing the period.
* `folds`: Number of k-folds to apply.
* `n_valPeriods`: Periods in the validation set (assumed to be the last ones).
* `orderPeriods`: Chronologically ordered vector of periods.

```{r}
# Define a dataset with 36 periods (monthly) and one observation per period
# There could be more than one observation per period.
set.seed(273)
years <- c(2023, 2024, 2025)
total_periods <- c(sapply(years, function(y){
  months <- 1:12
  months[1:9] <- paste0("0", months[1:9])
  return(paste0("MM", months, y))
}, USE.NAMES = FALSE))
target <- rnorm(n = length(total_periods), mean = 10, sd = 1)
cv_t_sampler_data.dt <- data.table::data.table(measure = target, period = total_periods)
head(cv_t_sampler_data.dt)

# Define the sampler
cv_t_sampler <- cvTemporalSampler(key = 1:nrow(cv_t_sampler_data.dt), # keys
                                  data = cv_t_sampler_data.dt, # data
                                  identifier = "period", # Name of the column with periods
                                  folds = 3, # 3 partitions
                                  n_valPeriods = 6, # We require 2 months for validation
                                  orderPeriods = total_periods) # Ordered periods

# As can be seen, the third sample would take the last 6 months as validation,
# the second would take 12 and the first 18.
getSample(cv_t_sampler)
```



